
title: "GEDCOM to Pedigree Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidygedcom)
```



Once the data is in a tidy data frame, the `BGmisc` package provides the next layer of analysis: computing pairwise relatedness, tracing paternal lineages, and identifying Y-chromosome carriers.

The strategy for the Waugh paternity question is:

1. Identify the focal individual and trace his Y-chromosome line back to William Pitt Waugh Sr.
2. Flag all individuals descended from William Pitt Waugh Sr. who are **not** in W. Henderson Waugh's branch — these are descendants of William Pitt Waugh Jr. (the Segraves line).
3. Among those, find living males: they carry the same Y chromosome as William Pitt Waugh Sr. and are potential DNA match candidates.
4. A Y-chromosome match between a living Segraves-line male and the focal individual (or his close Y-line relatives) would strongly confirm that W. Henderson Waugh and William Pitt Waugh Jr. share the same father.

```{r bgmisc-workflow, eval = requireNamespace("BGmisc", quietly = TRUE) && requireNamespace("ggpedigree", quietly = TRUE)}
library(BGmisc)
library(ggpedigree)

# Convert IDs to numeric (required by BGmisc)
ped_num <- ped
ped_num$personID <- as.numeric(ped_num$personID)
ped_num$momID    <- as.numeric(ped_num$momID)
ped_num$dadID    <- as.numeric(ped_num$dadID)

# Structure pedigree into family units and compute paternal line IDs
ged_ped <- ped2fam(ped_num, personID = "personID") |>
  ped2paternal(personID = "personID", momID = "momID", dadID = "dadID")

# Identify John William "Bud" Waugh as the focal person's great-grandfather
# (the deepest ancestor in our sample with a known paternal chain)
focal_ID  <- ged_ped$personID[grepl("John William", ged_ped$name)]
y_line_ID <- ged_ped$patID[ged_ped$personID == focal_ID]

# Flag Y-line membership — green branch = focal person's line
ged_ped$y_line <- ifelse(ged_ped$patID == y_line_ID, "Y Line", "Other Lines")
```

```{r pedigree-plot, eval = requireNamespace("ggpedigree", quietly = TRUE), fig.width = 7, fig.height = 4, fig.alt = "Pedigree plot of the Waugh family sample showing William Pitt Waugh Sr. at top with two branches: the focal person's Y line through W. Henderson Waugh on the left, and the Segraves Y-DNA candidate branch through William Pitt Waugh Jr. on the right."}
ggpedigree(
  ged_ped,
  personID = "personID",
  momID    = "momID",
  dadID    = "dadID",
  sexVar   = "sex",
  config   = list(
    label_include  = TRUE,
    label_column   = "name",
    label_text_size = 2,
    code_male = "M",
    code_female = "F",
    segment_lineage_include          = TRUE,
    segment_lineage_focal_personID   = focal_ID,
    segment_lineage_component        = "paternal",
    segment_lineage_legend_title     = "Patriline",
  add_phantoms = TRUE,
  founder_order_seed = 1L
  )
)
```

The plot shows both branches descending from William Pitt Waugh Sr.: the green (focal) line through W. Henderson Waugh on the left, and the Segraves branch through William Pitt Waugh Jr. on the right. These are distinct people with distinct IDs — the shared surname is what makes the tree appear circular at first glance, but there is no loop.

In the full Waugh analysis, `addPaternalChain()` and `addPaternalLineFlag()` from `BGmisc` extend this further by flagging descendants of specific ancestors. The key variable is:

```r
in_granddad_henderson_line_not_dad_henderson_line =
  in_granddad_henderson_line & !in_dad_henderson_line
```

This isolates individuals descended from William Pitt Waugh Sr. but **not** through W. Henderson Waugh — i.e., the Segraves (William Pitt Jr.) branch — who are the primary Y-DNA match candidates. `ggPedigreeInteractive()` renders the full tree with nodes colored by additive genetic relatedness to the focal individual and edges highlighted by patriline membership.



## Working with a real GEDCOM file

To apply this workflow to a real Ancestry.com export:

```{r real-file, eval = FALSE}
# Unzip the Ancestry export
unzip("W. Henderson Waugh Family Tree.zip", overwrite = TRUE)

# Peek at the raw file to orient yourself before parsing
raw_ged <- readLines("W. Henderson Waugh Family Tree.ged")
head(raw_ged, 15)

# Locate the anchor individual by name
line_num <- which(grepl("1 NAME W. Henderson /Waugh/", raw_ged, fixed = TRUE))
raw_ged[line_num:(line_num + 20)]

# Parse all INDI blocks
ped <- readGedcom("W. Henderson Waugh Family Tree.ged",
                  remove_empty_cols = TRUE, verbose = FALSE)

# Add year columns robust to approximate dates
ped$birth_year <- extractGedcomYear(ped$birth_date)
ped$death_year <- extractGedcomYear(ped$death_date)

# Quick overview
summarizeGedcom(ped)

# Parse family/marriage records
fam <- readGedcomFamilies("W. Henderson Waugh Family Tree.ged", verbose = FALSE)
```

Note that Ancestry.com regenerates person IDs each time you export, so ID values will differ between exports. Always locate individuals by name or birth date rather than by a hardcoded ID.



## Cleaning up

```{r cleanup}
unlink(tmp_ged)
```



## Summary of functions

| Function | Purpose |
|||
| `readGedcom(file)` | Parse `INDI` blocks → one row per person |
| `readGedcomFamilies(file)` | Parse `FAM` blocks → one row per family |
| `summarizeGedcom(df)` | Coverage counts and percentages |
| `extractGedcomYear(x)` | Year from any GEDCOM date string |
| `convertGedcomCoords(df)` | Convert `_lat`/`_long` columns to decimal degrees |
| `gedcomLatToNumeric(x)` | Convert a latitude string vector |
| `gedcomLonToNumeric(x)` | Convert a longitude string vector |
